A non-contextual hidden variable model for quantum mechanics

### 2016-07-13 finalPaper.tex
some small change in finalPaper.tex :P
removed the footnote etc.

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%This stuff is required for transposing matrices | fine I didn't want to do donkey work
% \usepackage{xparse}
% \usepackage{environ}

% \ExplSyntaxOn
% \NewEnviron{bmatrixT}
%  {
%   \marine_transpose:V \BODY
%  }

% \int_new:N \l_marine_transpose_row_int
% \int_new:N \l_marine_transpose_col_int
% \seq_new:N \l_marine_transpose_rows_seq
% \seq_new:N \l_marine_transpose_arow_seq
% \prop_new:N \l_marine_transpose_matrix_prop
% \tl_new:N \l_marine_transpose_last_tl
% \tl_new:N \l_marine_transpose_body_tl

% \cs_new_protected:Nn \marine_transpose:n
%  {
%   \seq_set_split:Nnn \l_marine_transpose_rows_seq { \\ } { #1 }
%   \int_zero:N \l_marine_transpose_row_int
%   \prop_clear:N \l_marine_transpose_matrix_prop
%   \seq_map_inline:Nn \l_marine_transpose_rows_seq
%    {
%     \int_incr:N \l_marine_transpose_row_int
%     \int_zero:N \l_marine_transpose_col_int
%     \seq_set_split:Nnn \l_marine_transpose_arow_seq { & } { ##1 }
%     \seq_map_inline:Nn \l_marine_transpose_arow_seq
%      {
%       \int_incr:N \l_marine_transpose_col_int
%       \prop_put:Nxn \l_marine_transpose_matrix_prop
%        {
%         \int_to_arabic:n { \l_marine_transpose_row_int }
%         ,
%         \int_to_arabic:n { \l_marine_transpose_col_int }
%        }
%        { ####1 }
%      }
%    }
%    \tl_clear:N \l_marine_transpose_body_tl
%    \int_step_inline:nnnn { 1 } { 1 } { \l_marine_transpose_col_int }
%     {
%      \int_step_inline:nnnn { 1 } { 1 } { \l_marine_transpose_row_int }
%       {
%        \tl_put_right:Nx \l_marine_transpose_body_tl
%         {
%          \prop_item:Nn \l_marine_transpose_matrix_prop { ####1,##1 }
%          \int_compare:nF { ####1 = \l_marine_transpose_row_int } { & }
%         }
%       }
%      \tl_put_right:Nn \l_marine_transpose_body_tl { \\ }
%     }
%    \begin{bmatrix}
%    \l_marine_transpose_body_tl
%    \end{bmatrix}
%  }
% \cs_generate_variant:Nn \marine_transpose:n { V }
% \cs_generate_variant:Nn \prop_put:Nnn { Nx }
% \ExplSyntaxOff

\begin{document}
\title{A non-contextual hidden variable model for quantum mechanics}
\author{Atul Singh Arora}
%\email{ms11003@iisermohali.ac.in}
\selectlanguage{british}%
\affiliation{Indian Institute of Science Education and Research (IISER), Mohali}
\author{Arvind}
%\email{arvind@iisermohali.ac.in}
\selectlanguage{british}%
\affiliation{Indian Institute of Science Education
and Research (IISER), Mohali}
\begin{abstract}
We propose a non-contextual hidden variable model,
consistent with all predictions of quantum
mechanics (QM).  A careful scrutiny of consistency
requirements of any hidden variable model with
quantum mechanics and the corresponding no-go
theorems, leads us to the conclusion that the
notion of contextuality is not a necessary feature
of QM.  The alternative view that emerges, hinges
on identifying a new classical property, which we
call ``multiplicativity''.  It turns out that by
relaxing this condition of ``multiplicativity'' on
hidden variable models, they can be made
consistent with QM.  The connection of
non-multiplicativity to Bohm's model and to
apparent contextuality is also discussed.
\end{abstract}
\pacs{}
\maketitle
Quantum mechanics has been one of the most
successful theories in physics so far, however,
there has not yet been a final word on its
completeness and interpretation.
Einstein's~\cite{EinsteinEPR} work on the
incompleteness of QM and the subsequent seminal
work of Bell~\cite{Bell1964}, assessing the
compatibility of a more complete model involving
hidden variables (HV) and locality with QM, has
provided deep insights into  how the quantum world
differs from its classical counterpart.  In recent
times, these insights have been of pragmatic
utility in the areas of quantum information
processing, where EPR pairs are fundamental motifs
of entanglement~\cite{Ekert,PironioRndmnssCrtfcn,NielsenChuang}.
The work of  Kochen Specker~\cite{KochenSpecker}
and Gleason et.
al.~\cite{Gleason,BellOnHiddenVariables,Peres,Mermin})
broadened the schism between  HV models and QM.
They showed that it was contextuality and not
non-locality which was at the heart of this schism
and the incompatibility between HV models and QM
can arise even for a single quantum system.
Contextuality has thus been identified as a
fundamental non-classical feature of the quantum
world and experiments have also been proposed and
conducted to this effect~\cite{SimonContExpProp,
HuangContExp}.
Contextuality, on the one hand has led to
investigations on the foundational aspects of
QM including attempts to prove completeness of QM~\cite{PawelCntxClsscl,CabelloMmryQM}, and on the other
hand has been harnessed for computation and
cryptography~\cite{HowardCntxCmptn,CabelloCntxScrt}.
However, not all HV
models are incompatible with QM, and a very
important case in point is  Bohm's model based on
precise trajectories for quantum
systems~\cite{Bohm1,Bohm2}.

@@

%{\noindent \bf Notation:} 
\begin{notn*}
(a) $\psi\in\mathcal{H}$ 
represents a pure quantum mechanical
state of the system in the Hilbert space
$\mathcal{H}$, (b) $\hat{\mathcal{H}}$ is defined
to mean the set of Hermitian 
observables for the system, (c)
$[\mathcal{H}]$ is defined to mean
$(\mathcal{H},\,\mathbb{R}^{\otimes})$, which
represents the state of the system including HVs,
(d) $[\psi]\in[\mathcal{H}]$ will represent the
state of the system, including HVs, (e) a
prediction map is
$M:\{ \hat{\mathcal{H}},[\mathcal{H}] \}\to\mathbb{R}$,
(f) a sequence map is
$S:\{ \hat{\mathcal{H}},[\mathcal{H}],\mathbb{R} \}\to[\mathcal{H}]$,
(g) $f$ is an arbitrary map from $\{
\hat{\mathcal{H}},\hat{\mathcal{H}},\dots\hat{\mathcal{H}}
\} \to \hat{\mathcal{H}}$ constructed using
multiplication and addition of the observables,
and multiplication with complex
numbers\footnote{Strictly, the map $f$ would
depend on the operators to ensure preservation of
Hermiticity; thus a given $f$ will be defined only
for a subset of $\{
\hat{\mathcal{H}},\hat{\mathcal{H}},\dots\hat{\mathcal{H}}
\}$.}, (h) $\tilde{f}$ is a map constructed by
replacing observables in $f$ with real numbers.
\end{notn*}

@@

\begin{defnrem} The notion of non-contextuality may be
extended to maps. A map is non-contextual if it is
of the form $M: \{ \hat{\mathcal{H}},[\mathcal{H}]
\} \to\mathbb{R}$; prediction maps as defined above, for individual observables, are non-contextual by construction.  
\end{defnrem}
\begin{defnrem}
Broader definitions in the literature have been
suggested which declare a larger set of theories
as non-contextual. For our purposes however, this
restricted definition will suffice. The term
contextual is used to suggest that the value an
operator takes might depend on which other
compatible observable it is being  measured with.
\end{defnrem}

@@

Before proceeding we note, however, that QM can
enforce \emph{sequential multiplicativity}. This
result will be used to assess accuracy of any
proposal of a HV
model.
\begin{prop*} 
Let a quantum mechanical 
system be in a state, s.t. measurement of
$\hat{C}$ yields repeatable results (same result
each time). Then according to QM, \emph{sequential
multiplicativity} holds, where $\hat{C}:=
f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{n})$, and
$\hat{B}_{i}$ are as defined (in \defnref{seqnMltpl})
\end{prop*}
\begin{proof} 
Without loss of generality we can take
$\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{n}$ to be
mutually compatible and a complete set of
observables (operators can be added to make the set
complete if it not so).
It follows
that $\exists$
$\left|\mathbf{b}=\left(b_{1},b_{2},\dots
b_{n}\right)\right\rangle $ s.t.
$\hat{B}_{i}\left|\mathbf{b}\right\rangle
=b_{i}\left|\mathbf{b}\right\rangle $,
and that
$\sum_{\mathbf{b}}\left|\mathbf{b}\right\rangle
\left\langle \mathbf{b}\right|=\hat{\mathbb{I}}$.
Let the state of the system
$\left|\psi\right\rangle $ be s.t.
$\hat{C}\left|\psi\right\rangle
=c\left|\psi\right\rangle $. For
the statement to follow, one need only show that
$\left|\psi\right\rangle $ must be made of only
those $\left|\mathbf{b}\right\rangle $s, which
satisfy $c=\tilde
f(b_{1},b_{2},\dots
b_{n})$.  This is the crucial step and
proving this is straightforward. We start with
$\hat{C}\left|\psi\right\rangle
=c\left|\psi\right\rangle $ and take its inner
product with $\left\langle \mathbf{b}\right|$ to
get \begin{eqnarray*} \left\langle
\mathbf{b}\right|\hat{C}\left|\psi\right\rangle  &
= & c\left\langle \mathbf{b}|\psi\right\rangle ,\\
\left\langle
\mathbf{b}\right|f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{n})
\left|\psi\right\rangle
& = & c\left\langle \mathbf{b}|\psi\right\rangle
,\\ \tilde f(b_{1},b_{2},\dots
b_{n})\left\langle
\mathbf{b}|\psi\right\rangle  & = & c\left\langle
\mathbf{b}|\psi\right\rangle .  \end{eqnarray*}
Also, we have $\left|\psi\right\rangle
=\sum_{\mathbf{b}}\left\langle
\mathbf{b}|\psi\right\rangle
\left|\mathbf{b}\right\rangle $, from
completeness. If we consider
$\left|\mathbf{b}\right\rangle $s for which
$\left\langle \mathbf{b}|\psi\right\rangle \neq0$,
then we can conclude that indeed $c=\tilde
f(b_{1},b_{2},\dots
b_{n})$. However, when $\left\langle
\mathbf{b}|\psi\right\rangle =0$, viz.
$\left|\mathbf{b}\right\rangle $s that are
orthogonal to $\left|\psi\right\rangle $, then
nothing can be said but that is irrelevant. We can thus conclude that
$\left|\psi\right\rangle $ is made only of those
$\left|\mathbf{b}\right\rangle $s that satisfy the
required relation.
\end{proof} 
It is worth noting that in the Peres Mermin
case, where $\hat{R}_{i}$ and $\hat{C}_{j}$ are
just $\pm\hat{\mathbb{I}}$, it follows that all
states are their eigenstates. Consequently, for
these operators \emph{sequential multiplicativity}
must always hold.
%\subsection{C-ingle model | An Explicit
%Construction \label{ssec:cingle}}
We are now ready to describe our explict model.
Let the state of a finite dimensional quantum system
be $\left|\chi\right\rangle $. We wish to
assign a value to an arbitrary observable 
$\hat{A}=\sum_{a}a\left|a\right\rangle
\left\langle a\right|$, which has 
eigenvectors $\{ \vert
a_j\rangle \}$. The corresponding ordered eigenvalues are $\{a_j\}$ such that 
$a_{\rm min}=a_1$
and 
$a_{\rm max}=a_n$.

@@

Note that the assignment described by the
prediction map $M$ is non-contextual since, given
an operator and a state (+ the HV $c$), the value is
uniquely assigned. The map $M$ is however,
non-multiplicative.

To bring out the non-multiplicativity of our model 
and to convince the reader of its generic validity,
we apply it to the Peres Mermin situation of two
spin-half particles. Consider the initial state of
the system 
$\left|\psi_{1}\right\rangle
=\left|00\right\rangle $. Assume we 
obtained $c=0.4$ as a random choice. To arrive
at the assignments, note that
$\left|00\right\rangle $ is an eigenket of only
$\hat{R}_{i},\hat{C}_{j}$ and
$\hat{A}_{33}=\hat{\sigma}_{z}\otimes\hat{\sigma}_{z}$.
Thus, in the first iteration, all these should be
assigned their respective eigenvalues. The
remaining operators must be assigned $-1$ as one
can readily verify by explicitly finding the
smallest $a$ as described in postulate 2 of the
model (see \tblref{HVmodel}). 
%Two remarks are in
This model is manifestly
\emph{non-multiplicative}, for
$M_{1}(\hat{C}_{3})=1\neq
M_{1}(\hat{A}_{13})M_{1}(\hat{A}_{23})M_{1}(\hat{A}_{33})=-1$,
where the subscript refers to the iteration
number.
More precisely,
$M_{1}(\hat{O}):=M(\hat{O},\left[\left|\psi_{1}\right\rangle
=\left|00\right\rangle \right])$ where the
complete state $\left[\left|\psi_{1}\right\rangle
\right]$ implicitly refers to both the quantum
state $\left|00\right\rangle $ and the HV $c=0.4$.
%Second, we must impose \emph{sequential
%multiplicativity} as a consistency check of the
%model, which in particular entails that
The model however  obeys the \emph{sequential
multiplicativity} as required for consistency,
namely,
$M_{1}(\hat{C}_{3})=M_{1}(\hat{A}_{33})M_{2}(\hat{A}_{23})M_{3}(\hat{A}_{13})$,
where
$M_{2}:=M(\hat{O},\left[\left|\psi_{2}\right\rangle
\right])$,
$M_{3}:=M(\hat{O},\left[\left|\psi_{3}\right\rangle
\right])$ and $\left|\psi_{2}\right\rangle
,\,\left|\psi_{3}\right\rangle $ are obtained from
postulate 3. Note that for each iteration, a new
HV is generated. To illustrate \emph{sequential
multiplicativity}, we must choose to measure
$\hat{A}_{33}$. According to step 3, since
$\left|00\right\rangle $ is an eigenstate of
$\hat{A}_{33}$, the final state remains
$\left|00\right\rangle $.

@@

% \text{(Assignments)} &
% m_{1}(\hat{R}_{i}),m_{1}(\hat{C}_{j})=+1\,(j\neq3)
% &
% m_{2}(\hat{R}_{i}),m_{2}(\hat{C}_{j})=+1\,(j\neq3)
% &
% m_{3}(\hat{R}_{i}),m_{3}(\hat{C}_{j})=+1\,(j\neq3)\\
% & m_{1}(\hat{C}_{3})=-1 & m_{2}(\hat{C}_{3})=-1 &
% m_{3}(\hat{C}_{3})=-1
% \end{bmatrixT}
\label{eq:toyModel}
\end{array}
% \label{eq:toyModel}
\end{equation*}
\caption{HV model applied to the Peres Mermin situation}
\label{tbl:HVmodel}
\end{table*}
This has implications to Bohm's HV model (BHVM),
where we have the notion of precise trajectories
for particles guided by the wavefunction.  In BHVM
if the initial conditions are precisely known the
entire trajectory of a particle can be predicted
including the individual outcome of measurements.
For the case of a single spin-half particle in a
Stern-Gerlach experiment, BHVM can predict
opposite results for two measurements of the
same operator.  This observation is often used as
a demonstration of `contextuality' in BM, although
we did not even speak of compatible
measurements~\cite{HardyCntxBM}.  It turns out, however,
that if one restricts oneself to a one-one map
between an experiment and an observable (by
following a certain convention), the so-called
`contextuality' can be removed from BHVM.  One
finds that the map so obtained from observables to
their predictions turns out to be
non-multiplicative, suggesting that at least in
this case, non-multiplicativity is a more suitable
explanation of non-classicality.

@@

Imagine
$
\hat{B}_{1}=\hat{\sigma}_{z}\otimes\hat{\mathbb{I}}=
\left|00\right\rangle
\left\langle 00\right|+\left|01\right\rangle
\left\langle 01\right|-\left[\left|10\right\rangle
\left\langle 10\right|+\left|11\right\rangle
\left\langle 11\right|\right],
$ 
$
\hat{B}_{2}=\hat{\mathbb{I}}\otimes\hat{\sigma}_{z}=
\left|10\right\rangle
\left\langle 10\right|+\left|11\right\rangle
\left\langle 11\right|-\left[\left|00\right\rangle
\left\langle 00\right|+\left|01\right\rangle
\left\langle 01\right|\right],
$ 
while we define 
$
\hat{C}=f(\{\hat{B}_{i}\})
=0.\left|00\right\rangle \left\langle
00\right|+1.\left|01\right\rangle \left\langle
01\right|+2.\left|10\right\rangle \left\langle
10\right|+3.\left|11\right\rangle \left\langle
11\right|. 
$
$\hat{C}$ maybe viewed as a function of
$\hat{B}_{1}$, $\hat{B}_{2}$ and other operators
$\hat{B}_{i}$ which are constructed to obtain a
maximally commuting set. A measurement of
$\hat{C}$, will collapse the state into one of the
states which are simultaneous eignkets of $B_{1}$
and $B_{2}$. Consequently, from the observed value
of $\hat{C}$, one can deduce the values of
$\hat{B}_{1}$ and $\hat{B}_{2}$. Now consider
$\sqrt{2}\left|\chi\right\rangle
=\left|10\right\rangle +\left|01\right\rangle $,
for which $M_{1}(\hat{B}_{1})=1$, and
$M_{1}(\hat{B}_{2})=1$, using our model,
with $c<0.5$. However, $M_{1}(\hat{C})=1$, from
which one can deduce that $B_{1}$ was $+1$, while
$B_{2}$ was $-1$.  This property itself, one may
be tempted call contextuality, viz.  the value of
$B_{1}$ depends on whether it is measured alone or
with the remaining $\{B_{i}\}$. However, it must
be noted that $B_{1}$ has a well defined value,
and so does $\hat{C}$. Thus by our accepted
definition, there is no contextuality. It is just
that $M_{1}(\hat{C})\neq \tilde
f(M_{1}(\hat{B}_{1}),M_{1}(\hat{B}_{2}),\dots)$,
viz. the theory is non-multiplicative. Note that
after measuring $\hat{C}$ however,
$M_{2}(\hat{B}_{1})=+1$ and
$M_{2}(\hat{B}_{2})=-1$ (for any value of $c$)
consistent with those deduced by measuring
$\hat{C}$.

Evidently, multiplicativity must hold for the
common eigenkets of $\hat{B}_{i}$.  Consequently,
any violation of multiplicativity must arise from
states that are superpositions or linear
combinations of these common eigenkets.  Note,
however, that entanglement is not necessary to
show a violation; for instance the Peres Mermin
test is a state independent test where a separable
state can be used to arrive at a contradiction.
Entanglement can be used to show non-locality of
non-multiplicative prediction maps using Bell's
theorem however, the notion of non-multiplicativity
is more basic.  \begin{acknowledgments} 
ASA and Arvind acknowledge funding from KVPY and DST Grant No. EMR/2014/000297 respectively.
%Arvind acknowledges funding from DST India under Grant No. EMR/2014/000297. ASA acknolwedges KVPY, DST for financial support.
\end{acknowledgments}
%\bibliographystyle{apsrev4-1}
\bibliography{references} \end{document}

>>>>

\begin{document}
\title{A non-contextual hidden variable model for quantum mechanics}
\author{Atul Singh Arora}
\email{ms11003@iisermohali.ac.in}
\selectlanguage{british}%
\affiliation{Indian Institute of Science Education and Research (IISER), Mohali}
\author{Arvind}
\email{arvind@iisermohali.ac.in}
\selectlanguage{british}%
\affiliation{Indian Institute of Science Education
and Research (IISER), Mohali}
\begin{abstract}
We propose a non-contextual hidden variable model,
consistent with all predictions of quantum
mechanics (QM).  A careful scrutiny of consistency
requirements of any hidden variable model with
quantum mechanics and the corresponding no-go
theorems, leads us to the conclusion that the
notion of contextuality is not a necessary feature
of QM.  The alternative view that emerges, hinges
on identifying a new classical property which we
call ``multiplicativity''.  It turns out that by
relaxing this condition of ``multiplicativity'' on
hidden variable models, they can be made
consistent with QM.  The connection of
non-multiplicativity to Bohm's model and to
apparent contextuality is also discussed.
\end{abstract}
\pacs{}
\maketitle
Quantum mechanics has been one of the most
successful theories in physics so far, however,
there has not yet been a final word on its
completeness and interpretation~\cite{BellSpkblUnspkbl}.
Einstein's~\cite{EinsteinEPR} work on the
incompleteness of QM and the subsequent seminal
work of Bell~\cite{BellSpkblUnspkbl}, assessing the
compatibility of a more complete model involving
hidden variables (HV) and locality with QM, has
provided deep insights into  how the quantum world
differs from its classical counterpart.  In recent
times, these insights have been of pragmatic
utility in the areas of quantum information
processing, where EPR pairs are fundamental motifs
of entanglement~\cite{Ekert,PironioRndmnssCrtfcn,NielsenChuang}.
The work of  Kochen Specker~\cite{KochenSpecker}
and Gleason et.
al.~\cite{Gleason,BellOnHiddenVariables,Peres,Mermin})
broadened the schism between  HV models and QM.
They showed that it was contextuality and not
non-locality which was at the heart of this schism
and the incompatibility between HV models and QM
can arise even for a single quantum system.
Contextuality has thus been identified as a
fundamental non-classical feature of the quantum
world and experiments have also been proposed and
conducted to this effect~\cite{SimonContExpProp,
HuangContExp}.
Contextuality, on the one hand has led to
investigations on the foundational aspects of
QM including attempts to prove completeness of QM~\cite{PawelCntxClsscl,CabelloMmryQM}, and on the other
hand has been harnessed for computation and
cryptography~\cite{HowardCntxCmptn,CabelloCntxScrt}.
However, not all HV
models are incompatible with QM, and a very
important case in point is  Bohm's model based on
precise trajectories for quantum
systems~\cite{Bohm1,Bohm2}.

@@

%{\noindent \bf Notation:} 
\begin{notn*}
(a) $\psi\in\mathcal{H}$ 
represents a pure quantum mechanical
state of the system in the Hilbert space
$\mathcal{H}$, (b) $\hat{\mathcal{H}}$ is defined
to mean the set of Hermitian 
observables for the system, (c)
$[\mathcal{H}]$ is defined to mean
$(\mathcal{H},\,\mathbb{R}^{\otimes})$, which
represents the state of the system including HVs,
(d) $[\psi]\in[\mathcal{H}]$ will represent the
state of the system, including HVs, (e) a
prediction map is
$M:\{ \hat{\mathcal{H}},[\mathcal{H}] \}\to\mathbb{R}$,
(f) a sequence map is
$S:\{ \hat{\mathcal{H}},[\mathcal{H}],\mathbb{R} \}\to[\mathcal{H}]$,
(g) $f$ is an arbitrary map from $\{
\hat{\mathcal{H}},\hat{\mathcal{H}},\dots\hat{\mathcal{H}}
\} \to \hat{\mathcal{H}}$ constructed using
multiplication and addition of compatible observables,
and multiplication with complex
numbers
%\footnote{Strictly, the map $f$ would depend on the operators to ensure preservation of Hermiticity; thus a given $f$ will be defined only for a subset of $\{\hat{\mathcal{H}},\hat{\mathcal{H}},\dots\hat{\mathcal{H}}\}$.}
, (h) $\tilde{f}$ is a map constructed by
replacing observables in $f$ with real numbers.
\end{notn*}

@@

\begin{defnrem} The notion of non-contextuality may be
extended to maps. A map is non-contextual if it is
of the form $M: \{ \hat{\mathcal{H}},[\mathcal{H}]
\} \to\mathbb{R}$; prediction maps as defined above, for individual observables, are non-contextual by construction.  
\end{defnrem}
\begin{defnrem}
Broader definitions in the literature have been
suggested which declare a larger set of theories
as non-contextual. For our purposes, however, this
restricted definition will suffice. The term
contextual is used to suggest that the value an
operator takes might depend on which other
compatible observable it is being  measured with.
\end{defnrem}

@@

Before proceeding we note, however, that QM can
enforce \emph{sequential multiplicativity}. This
result will be used to assess accuracy of any
proposal of a HV
model.
\begin{prop*} 
Let a quantum mechanical 
system be in a state, s.t. measurement of
$\hat{C}$ yields repeatable results (same result
each time). Then according to QM, \emph{sequential
multiplicativity} holds, where $\hat{C}:=
f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{n})$, and
$\hat{B}_{i}$ are as defined (in \defnref{seqnMltpl})
\end{prop*}
\begin{proof} 
Without loss of generality we can take
$\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{n}$ to be
mutually compatible and a complete set of
observables (operators can be added to make the set
complete if it not so).
It follows
that $\exists$
$\left|\mathbf{b}=\left(b_{1},b_{2},\dots
b_{n}\right)\right\rangle $ s.t.
$\hat{B}_{i}\left|\mathbf{b}\right\rangle
=b_{i}\left|\mathbf{b}\right\rangle $,
and that
$\sum_{\mathbf{b}}\left|\mathbf{b}\right\rangle
\left\langle \mathbf{b}\right|=\hat{\mathbb{I}}$.
Let the state of the system
$\left|\psi\right\rangle $ be s.t.
$\hat{C}\left|\psi\right\rangle
=c\left|\psi\right\rangle $. For
the statement to follow, one need only show that
$\left|\psi\right\rangle $ must be made of only
those $\left|\mathbf{b}\right\rangle $s, which
satisfy $c=\tilde
f(b_{1},b_{2},\dots
b_{n})$.  This is the crucial step and
proving this is straightforward. We start with
$\hat{C}\left|\psi\right\rangle
=c\left|\psi\right\rangle $ and take its inner
product with $\left\langle \mathbf{b}\right|$ to
get \begin{eqnarray*} \left\langle
\mathbf{b}\right|\hat{C}\left|\psi\right\rangle  &
= & c\left\langle \mathbf{b}|\psi\right\rangle ,\\
\left\langle
\mathbf{b}\right|f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{n})
\left|\psi\right\rangle
& = & c\left\langle \mathbf{b}|\psi\right\rangle
,\\ \tilde f(b_{1},b_{2},\dots
b_{n})\left\langle
\mathbf{b}|\psi\right\rangle  & = & c\left\langle
\mathbf{b}|\psi\right\rangle .  \end{eqnarray*}
Also, we have $\left|\psi\right\rangle
=\sum_{\mathbf{b}}\left\langle
\mathbf{b}|\psi\right\rangle
\left|\mathbf{b}\right\rangle $, from
completeness. If we consider
$\left|\mathbf{b}\right\rangle $s for which
$\left\langle \mathbf{b}|\psi\right\rangle \neq0$,
then we can conclude that indeed $c=\tilde
f(b_{1},b_{2},\dots
b_{n})$. However, when $\left\langle
\mathbf{b}|\psi\right\rangle =0$, viz.
$\left|\mathbf{b}\right\rangle $s that are
orthogonal to $\left|\psi\right\rangle $, then
nothing can be said but that is irrelevant. We can thus conclude that
$\left|\psi\right\rangle $ is made only of those
$\left|\mathbf{b}\right\rangle $s that satisfy the
required relation.
\end{proof} 
It is worth noting that in the Peres Mermin
case, where $\hat{R}_{i}$ and $\hat{C}_{j}$ are
just $\pm\hat{\mathbb{I}}$, it follows that all
states are their eigenstates. Consequently, for
these operators \emph{sequential multiplicativity}
must always hold.
%\subsection{C-ingle model | An Explicit
%Construction \label{ssec:cingle}}
We are now ready to describe our explicit model.
Let the state of a finite dimensional quantum system
be $\left|\chi\right\rangle $. We wish to
assign a value to an arbitrary observable 
$\hat{A}=\sum_{a}a\left|a\right\rangle
\left\langle a\right|$, which has 
eigenvectors $\{ \vert
a_j\rangle \}$. The corresponding ordered eigenvalues are $\{a_j\}$ such that 
$a_{\rm min}=a_1$
and 
$a_{\rm max}=a_n$.

@@

Note that the assignment described by the
prediction map $M$ is non-contextual since, given
an operator and a state (+ the HV $c$), the value is
uniquely assigned. The map $M$ is, however,
non-multiplicative.

To bring out the non-multiplicativity of our model 
and to convince the reader of its generic validity,
we apply it to the Peres Mermin situation of two
spin-half particles. Consider the initial state of
the system 
$\left|\psi_{1}\right\rangle
=\left|00\right\rangle $. Assume we 
obtained $c=0.4$ as a random choice. To arrive
at the assignments, note that
$\left|00\right\rangle $ is an eigenket of only
$\hat{R}_{i},\hat{C}_{j}$ and
$\hat{A}_{33}=\hat{\sigma}_{z}\otimes\hat{\sigma}_{z}$.
Thus, in the first iteration, all these should be
assigned their respective eigenvalues. The
remaining operators must be assigned $-1$ as one
can readily verify by explicitly finding the
smallest $a$ as described in postulate 2 of the
model (see \tblref{HVmodel}). 
%Two remarks are in
This model is manifestly
\emph{non-multiplicative}, for
$M_{1}(\hat{C}_{3})=1\neq
M_{1}(\hat{A}_{13})M_{1}(\hat{A}_{23})M_{1}(\hat{A}_{33})=-1$,
where the subscript refers to the iteration
number.
More precisely,
$M_{1}(\hat{O}):=M(\hat{O},\left[\left|\psi_{1}\right\rangle
=\left|00\right\rangle \right])$ where the
complete state $\left[\left|\psi_{1}\right\rangle
\right]$ implicitly refers to both the quantum
state $\left|00\right\rangle $ and the HV $c=0.4$.
%Second, we must impose \emph{sequential
%multiplicativity} as a consistency check of the
%model, which in particular entails that
The model, however, obeys the \emph{sequential
multiplicativity} as required for consistency,
namely,
$M_{1}(\hat{C}_{3})=M_{1}(\hat{A}_{33})M_{2}(\hat{A}_{23})M_{3}(\hat{A}_{13})$,
where
$M_{2}:=M(\hat{O},\left[\left|\psi_{2}\right\rangle
\right])$,
$M_{3}:=M(\hat{O},\left[\left|\psi_{3}\right\rangle
\right])$ and $\left|\psi_{2}\right\rangle
,\,\left|\psi_{3}\right\rangle $ are obtained from
postulate 3. Note that for each iteration, a new
HV is generated. To illustrate \emph{sequential
multiplicativity}, we must choose to measure
$\hat{A}_{33}$. According to step 3, since
$\left|00\right\rangle $ is an eigenstate of
$\hat{A}_{33}$, the final state remains
$\left|00\right\rangle $.

@@

% \text{(Assignments)} &
% m_{1}(\hat{R}_{i}),m_{1}(\hat{C}_{j})=+1\,(j\neq3)
% &
% m_{2}(\hat{R}_{i}),m_{2}(\hat{C}_{j})=+1\,(j\neq3)
% &
% m_{3}(\hat{R}_{i}),m_{3}(\hat{C}_{j})=+1\,(j\neq3)\\
% & m_{1}(\hat{C}_{3})=-1 & m_{2}(\hat{C}_{3})=-1 &
% m_{3}(\hat{C}_{3})=-1
% \end{bmatrixT}
\label{eq:toyModel}
\end{array}
% \label{eq:toyModel}
\end{equation*}
\caption{HV model applied to the Peres Mermin situation}
\label{tbl:HVmodel}
\end{table*}
This has implications to Bohm's HV model (BHVM),
where we have the notion of precise trajectories
for particles guided by the wave-function.  In BHVM
if the initial conditions are precisely known the
entire trajectory of a particle can be predicted
including the individual outcome of measurements.
For the case of a single spin-half particle in a
Stern-Gerlach experiment, BHVM can predict
opposite results for two measurements of the
same operator.  This observation is often used as
a demonstration of `contextuality' in BM, although
we did not even speak of compatible
measurements~\cite{HardyCntxBM}.  It turns out, however,
that if one restricts oneself to a one-one map
between an experiment and an observable (by
following a certain convention), the so-called
`contextuality' can be removed from BHVM.  One
finds that the map so obtained from observables to
their predictions turns out to be
non-multiplicative, suggesting that at least in
this case, non-multiplicativity is a more suitable
explanation of non-classicality.

@@

Imagine
$
\hat{B}_{1}=\hat{\sigma}_{z}\otimes\hat{\mathbb{I}}=
\left|00\right\rangle
\left\langle 00\right|+\left|01\right\rangle
\left\langle 01\right|-\left[\left|10\right\rangle
\left\langle 10\right|+\left|11\right\rangle
\left\langle 11\right|\right],
$ 
$
\hat{B}_{2}=\hat{\mathbb{I}}\otimes\hat{\sigma}_{z}=
\left|10\right\rangle
\left\langle 10\right|+\left|11\right\rangle
\left\langle 11\right|-\left[\left|00\right\rangle
\left\langle 00\right|+\left|01\right\rangle
\left\langle 01\right|\right],
$ 
while we define 
$
\hat{C}=f(\{\hat{B}_{i}\})
=0.\left|00\right\rangle \left\langle
00\right|+1.\left|01\right\rangle \left\langle
01\right|+2.\left|10\right\rangle \left\langle
10\right|+3.\left|11\right\rangle \left\langle
11\right|. 
$
$\hat{C}$ maybe viewed as a function of
$\hat{B}_{1}$, $\hat{B}_{2}$ and other operators
$\hat{B}_{i}$ which are constructed to obtain a
maximally commuting set. A measurement of
$\hat{C}$, will collapse the state into one of the
states which are simultaneous eignkets of $\hat{B}_{1}$
and $\hat{B}_{2}$. Consequently, from the observed value
of $\hat{C}$, one can deduce the values of
$\hat{B}_{1}$ and $\hat{B}_{2}$. Now consider
$\sqrt{2}\left|\chi\right\rangle
=\left|10\right\rangle +\left|01\right\rangle $,
for which $M_{1}(\hat{B}_{1})=1$, and
$M_{1}(\hat{B}_{2})=1$, using our model,
with $c<0.5$. However, $M_{1}(\hat{C})=1$, from
which one can deduce that $\hat{B}_{1}$ was $+1$, while
$\hat{B}_{2}$ was $-1$.  This property itself, one may
be tempted call contextuality, viz.  the value of
$\hat{B}_{1}$ depends on whether it is measured alone or
with the remaining $\{\hat{B}_{i}\}$. However, it must
be noted that $\hat{B}_{1}$ has a well defined value,
and so does $\hat{C}$. Thus by our accepted
definition, there is no contextuality. It is just
that $M_{1}(\hat{C})\neq \tilde
f(M_{1}(\hat{B}_{1}),M_{1}(\hat{B}_{2}),\dots)$,
viz. the theory is non-multiplicative. Note that
after measuring $\hat{C}$, however,
$M_{2}(\hat{B}_{1})=+1$ and
$M_{2}(\hat{B}_{2})=-1$ (for any value of $c$)
consistent with those deduced by measuring
$\hat{C}$.

Evidently, multiplicativity must hold for the
common eigenkets of $\hat{B}_{i}$.  Consequently,
any violation of multiplicativity must arise from
states that are super-positions or linear
combinations of these common eigenkets.  Note,
however, that entanglement is not necessary to
show a violation; for instance the Peres Mermin
test is a state independent test where a separable
state can be used to arrive at a contradiction.
Entanglement can be used to show non-locality of
non-multiplicative prediction maps using Bell's
theorem, however, the notion of non-multiplicativity
is more basic.  \begin{acknowledgments} 
ASA and Arvind acknowledge funding from KVPY and DST Grant No. EMR/2014/000297 respectively.
%Arvind acknowledges funding from DST India under Grant No. EMR/2014/000297. ASA acknolwedges KVPY, DST for financial support.
\end{acknowledgments}
%\bibliographystyle{apsrev4-1}
\bibliography{references} \end{document}
